---
title: "R code for calculating EC~50~ and NEC using the kinetic BLM"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.	Load the packages

```{r message=FALSE, warning=FALSE}
library(deSolve) # for using the function "ode" to integrate
library(minpack.lm) # for using the function "nls.lm" to fit
library(ggplot2) # for plotting
library(ggrepel) # for adding text onto plots without overlapping
```


## 2.	BLM parameters estimated from this study

```{r}
K_Cd <- 10^7.96 # L/mol, stability constant for Cd2+ + BL = CdBL
K_Ca <- 10^4.34 # L/mol, stability constant for Ca2+ + BL = CaBL
K_Mg <- 10^4.21 # L/mol, stability constant for Mg2+ + BL = MgBL
K_Na <- 10^3.42 # L/mol, stability constant for Na+ + BL = NaBL
K_K  <- 10^4.01 # L/mol, stability constant for K+ + BL = KBL
K_H  <- 10^8.79 # L/mol, stability constant for H+ + BL = HBL
```


## 3.	TK-TD parameters estimated from this study

```{r}
ke <- 0.00308  # 1/hour, elimination rate constant of Cd
Jmax <- 3.16  # μg/g/h, maximum uptake rate of Cd 
CIT <- 33.5 # μg/g tissue, internal threshold concentration of Cd
kk <- 0.00267 # g tissue/ug Cd/hour, killing rate of Cd
h0 <- 0 # 1/hour, background hazard rate, set to 0 in this study 
```



## 4.	Define the kinetic BLM model  

```{r}
BLM <- function (t, y, parameters) {
  Cint <- y[1]
  Hazard <- y[2]
  C_Cd <- parameters[1]
  # μg Cd/g tissue/hour, uptake rate of Cd
  Jin <- Jmax*K_Cd*C_Cd/(1 +K_Cd*C_Cd +K_Ca*C_Ca +K_Mg*C_Mg +K_Na*C_Na +K_K*C_K +K_H*C_H)
  dCint <- Jin - ke * Cint
  dHazard <- kk * max(Cint - CIT, 0) + h0
  list(c(dCint, dHazard))
}
```


## 5.	Define the objective function for fitting EC~50~

```{r}
fn_EC50 <- function(parameters_to_fit){
  C_Cd <- parameters_to_fit[1]
  out.fit <- ode(y = initial,
                 times = times,
                 func = BLM,
                 parms = c(C_Cd))
  n <- length(times)
  Survivorship_final <- exp(-out.fit[,3])[n]
  
  # definition of EC50: final survivorship = 0.5
  obj <- abs(Survivorship_final - 0.5) *1000
  return(obj)
}

```


## 6.	Define the function for calculating No-effect concentration (NEC)

```{r}
fn_NEC <- function(C_Cd) {
  # NEC is calculated by solving the equation: 
  # Jin/ke = CIT (or CIT * ke - Jin = 0)
  CIT * ke - (Jmax*K_Cd*C_Cd/(1 +K_Cd*C_Cd +K_Ca*C_Ca +K_Mg*C_Mg +K_Na*C_Na +K_K*C_K +K_H*C_H))
}
```


## 7.	Define initials for starting the integration

```{r}
initial <- c(Cint = 0, Hazard = 0)
```



## 8.	Read in the water chemistry data
The csv file should be placed in the “working directory” (Run the code getwd() to get the location of “working directory”). The csv file can be downloaded from the link: https://github.com/tan-qiao-guo/kineticBLM.

```{r}
data_WaterChem <- read.csv("WaterChem_input.csv") 
```

Alternatively, the data can be directly imported by reading the csv file deposited on github:

```r
url <- "https://raw.githubusercontent.com/tan-qiao-guo/kineticBLM/master/WaterChem_input.csv"
data_WaterChem <- read.csv(url) 
```


## 9.	Calculate 48-h EC~50~ and NEC in test solutions of different water chemistry

```{r}
# Input the duration of exposure
duration <- 48  # hour, duration of exposure
times <- seq(from = 0, to = duration, by = 0.1)
parameters <- c(C_Cd = 1e-7)  # mol/L, initial value free ion activity of Cd2+
for (i in 1: length(data_WaterChem[,1])) {
  C_Ca <- data_WaterChem[i,3] #mol/L, free ion activity of Ca2+
  C_Mg <- data_WaterChem[i,4] #mol/L, free ion activity of Mg2+
  C_Na <- data_WaterChem[i,5] #mol/L, free ion activity of Na+
  C_K <- data_WaterChem[i,6] #mol/L, free ion activity of K+
  C_H <- data_WaterChem[i,7] #mol/L, free ion activity of H+
  frac_free_ion <- data_WaterChem[i,8] #ratio of free ion of Cd2+ to total Cd 
  act_coef <- data_WaterChem[i,9] # activity coefficient of Cd2+

  #calculate the EC50 based on input water chemistry 
  fit_EC50 <- optim(par=parameters, fn=fn_EC50,
                    lower=1e-9,
                    upper=1e-3,
                    method="Brent")
  
  # Converte EC50 from free ion activity (mol/L) to concentration (μg/L) 
  EC50 <- as.numeric(fit_EC50$par)/frac_free_ion/act_coef * 112.4 * 1e6
  
  # Add the estimated EC50 into the table
  data_WaterChem$EC50[i] <- EC50
  
  # Fit NEC 
  fit_NEC <- uniroot(f = fn_NEC,
                     interval = c(1e-12, 1e-6),
                     tol = 1e-20)
  
  # Convert NEC from free ion activity (mol/L) to concentration (μg/L) 
  NEC <- fit_NEC$root/frac_free_ion/act_coef * 112.4 * 1e6
  
  # Add the estimated NEC into the table
  data_WaterChem$NEC[i] <- NEC
}

```


## 10.	Save the calculated EC50 and NEC values as an Excel file  

```{r}
write.csv(data_WaterChem,"output_EC50_NEC.csv")
```

The Excel file " output_EC50_NEC.csv " can be found in the working directory.


## 11.	Plot the EC~50~ and NEC 

```{r}
d <- subset(data_WaterChem, series !="Cd" & series != "Lakes")
# Select data other than the Cd series and the natural water series for plotting 

d$series <- factor(d$series,
                   levels=c("Ca","Mg","Na","K","pH","FA", "HA"))
# Reorder the data 
```


### 11.1 48-h EC50 plot
```{r}
p1 <- ggplot(d, aes(series, EC50, color=factor(level)))+
  theme_bw()+
  geom_jitter(position = position_jitter(width=0.05, height=0, seed=3),
              size=2.5)+
  scale_y_log10(limits=c(20, 2000))+
  scale_color_viridis_d(end=0.8, option="B")+
  geom_text_repel(aes(label=level),
                  size=3,
                  position = position_jitter(width=0.05, height=0, seed=3))+
  annotation_logticks(sides="l")+
  guides(color="none")+
  labs(x="Test series",
       y=expression("Cd 48-h"~EC[50]~"("*mu*g~L^"-1"*")"))+
  theme(axis.text.x = element_text(size=11, color="black"))

p1
```



### 11.2 No-effect concentration plot

```{r}
p2 <- ggplot(d, aes(series, NEC, color=factor(level)))+
  theme_bw()+
  geom_jitter(position = position_jitter(width=0.05, height=0, seed=2),
              size=2.5)+
  scale_y_log10(limits=c(1, 80))+
  scale_color_viridis_d(end=0.8)+
  geom_text_repel(aes(label=level),
                  size=3,
                  position = position_jitter(width=0.05, height=0, seed=2))+
  annotation_logticks(sides="l")+
  guides(color="none")+
  labs(x="Test series",
       y=expression("Cd no-effect concentration"~"("*mu*g~L^"-1"*")"))+
  theme(axis.text.x = element_text(size=11, color="black"))

p2
```


## 12.	Calculate Cd EC50 for different exposure time in Lake 6-Tingxi

```{r}
duration <- seq(40, 200, 1)  # hour, duration of exposure ranged from 40 h to 200 h
data_EC50vsTime <- data.frame(Time=duration, EC50=0) # create a table for storing calculated EC50
parameters <- 1e-7 # initial parameter value

for (i in 1: length(duration)) {
  times <- seq(from = 0, to = duration[i], by = 0.1)
  C_Ca <- data_WaterChem[57,3] #mol/L, free ion activity of Ca2+
  C_Mg <- data_WaterChem[57,4] #mol/L, free ion activity of Mg2+
  C_Na <- data_WaterChem[57,5] #mol/L, free ion activity of Na+
  C_K <- data_WaterChem[57,6] #mol/L, free ion activity of K+
  C_H <- data_WaterChem[57,7] #mol/L, free ion activity of H+
  frac_free_ion <- data_WaterChem[57,8] # ratio of free ion of Cd2+ to total Cd 
  act_coef <- data_WaterChem[57,9] # activity coefficient of Cd2+
  
  # fit the EC50
  fit_EC50 <- nls.lm(par=parameters, 
                     fn=fn_EC50,
                     lower=1e-9,
                     upper=1e-5,
                     control=nls.lm.control(maxiter = 500,
                                          ftol=1e-20,
                                          ptol = 1e-20))
  # Use the latest EC50 as the new intial parameter value
  parameters <- fit_EC50$par 
  
  # Convert EC50 from free ion activity (mol/L) to concentration (ug/L) 
  EC50 <- as.numeric(fit_EC50$par)/frac_free_ion/act_coef * 112.4 * 1e6
  
  # Add the estimated EC50 into the table
  data_EC50vsTime$EC50[i] <- EC50
}
```



## 13.	Plot Cd EC~50~ vs. exposure time

```{r}
ggplot(data_EC50vsTime, aes(Time, EC50))+
  theme_bw()+
  geom_line(color="navy")+
  scale_y_continuous(limits=c(0,40))+
  scale_x_continuous(breaks=seq(48,196,24))+
  labs(x="Exposure time (h)",
       y=expression("Cd"~EC[50]~"("*mu*g~L^"-1"*")"))
```




